---
title: "Bayesian Hidden Markov Models to Infer Latent Pitcher States"
author: "Garrett Lee, Trey Elder, Isaac Herman, Injun Lyo"
date: ""
output:
  pdf_document:
    extra_dependencies: ["float"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    fig.align = "center",
    fig.pos = "H"
)
```

# 1. Motivation

## What Problem Does This Method Solve?

Think about the last time you tried to figure out if a friend was in a bad mood. You couldn't see inside their head, but you could observe their behavior - maybe they snapped at a question, skipped lunch, or sighed loudly several times. From those observations, you made an inference: they're probably in a bad mood. That intuition - inferring something hidden from things you $can$ observe - is exactly what a Hidden Markov Model (HMM) does, just in a much more mathematical way.

To understand an HMM, it helps to first understand what a standard Markov model does. A Markov model is a probabilistic model describing a sequence of events with discrete states, where the probability of moving to the next state depends only on the current state - not on anything that came before it. As a simple example, suppose a friend named Bob does one of two activities each day: he either goes on a run or watches a movie, and tomorrow's activity depends only on today's activity. Each state has associate transition probabilities governing how likely he is to switch, If Bob ran today, there might be a 40% chance he runs tomorrow and a 60% chance he watches a movie. If he watched a movie today, there might be a 30% chance he watches again tomorrow and a 70% chance he runs.

```{r bob-markov, fig.align='center', fig.width=7, fig.height=4, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)

# Data for the 2x2 grid
grid_data <- data.frame(
  from  = c("Run", "Run", "Movie", "Movie"),
  to    = c("Run", "Movie", "Run", "Movie"),
  prob  = c(0.40, 0.60, 0.70, 0.30),
  label = c("40%", "60%", "70%", "30%")
)

# Set factor order so "Run" appears first (top/left)
grid_data$from <- factor(grid_data$from, levels = c("Run", "Movie"))
grid_data$to   <- factor(grid_data$to,   levels = c("Run", "Movie"))

ggplot(grid_data, aes(x = to, y = from)) +
  geom_tile(fill = "steelblue", alpha = 0.15, color = "steelblue", linewidth = 0.8) +
  geom_text(aes(label = label), size = 8, fontface = "bold", color = "steelblue") +
  scale_x_discrete(position = "top") +
  labs(
    title   = "Bob's Transition Probabilities",
    x       = "Tomorrow's Activity",
    y       = "Today's Activity"
  ) +
  theme_minimal() +
  theme(
    plot.title       = element_text(hjust = 0.5, size = 13, face = "bold", margin = margin(b = 10)),
    axis.title.x     = element_text(size = 11, face = "bold", margin = margin(b = 8)),
    axis.title.y     = element_text(size = 11, face = "bold", margin = margin(r = 8)),
    axis.text        = element_text(size = 11),
    panel.grid       = element_blank(),
    aspect.ratio     = 1
  )
```

A Hidden Markov Model extends this by introducing an underlying state that drives the observations - but that state is never directly observed. In Bob's case, suppose his daily activity is actually determined by the weather, which can either be rainy or sunny. The weather evolves as a Markov process, but since Bob lives far away, we never observe the weather directly. We only see what activity he does. That unobserved weather state, which generates the observable activity outcomes, is the core of an HMM.

Every HMM has three components:

1. **Hidden states** - a discrete set of underlying conditions that evolve over time according to a Markov process. In the Bob example, these are the two weather states, rainy and sunny.

2. **Transition matrix** - the probabilities of moving from one hidden state to another. For instance, if it's raining today, there might be a 60% chance of rain tomorrow and a 40% chance of sun. These transition probabilities define how the hidden process evolves.

3. **Emission probabilities** - given the current hidden state, the probability of each observable outcome. In our example, if it's sunny there's a 60% chance Bob runs and a 40% chance he watches a movie; if it's raining, those probabilities reverse to 30% and 70%. The observations we see are $emitted$ from the hidden state.

The key point is that we never observe the hidden states directly - we only ever see the emissions. The goal of the HMM is to work backwards: given the sequence of observations, what can we infer about the hidden states that generated them?

To make an HMM Bayesian, we place priors on the transition probabilities and emission probabilities, which are the parameters for the model. This allows us to incorporate domain knowledge or prior expectations into the model, and to quantify uncertainty. The Bayesian posterior we get also gives us full distributions over both model parameters and the hidden state sequences, rather than just point estimates.

## Why Can't Standard Regression Handle It?

There are three fundamental reasons why standard regression is not suited for HMM problems. First, the cause is invisible. Regression requires you to explicitly specify the variables driving the variation in your outcome. In an HMM setting, that driver is the hidden state - which, by definition, we do not observe. There is no covariate to include in a regression model, because the thing generating the outcomes is unobserved.

Second, the independence assumption is violated. Linear regression assumes that observations are independent (given the covariates). However is sequential HMM data, we essentially assume the opposite to be true. Each observation is dependent on the hidden state, and each hidden state is dependent on the previous hidden state. The encodes severe autocorrelation into our data. HMMs are explicitly designed to handle this temporal dependence, but linear regression is not.

Third, the data-generating process itself changes. Regression produces a single, fixed relationship between predictors and outcomes. In an HMM setting, that relationship shifts depending on which hidden state we are in. This is modeled in an HMM through different emission probability distributions for each hidden state. However, in linear regression, we would be forced to model this using two multiple regressions, one for each hidden state. However, we cannot do this because we don't know which observations belong to which state. HMMs solve this by estimating the hidden states and the model parameters simultaneuously.

## Where Does It Show Up in Practice?

Hidden Markov Models have a long history of real-world application across many fields. The original and most widely-known use is speech recognition, where the hidden states correspond to phonemes - the fundamental units of sound in a word - and the observations are raw acoustic signals. Before the era of deep learning and transformer models, HMMs were the backbone of speech recognition systems for decades.

In genomics, HMMs have been used to infer whether sequences of DNA correspond to coding or non-coding regions, based on pattersn in the nucleotide sequence. In finance, they are used to identify market regimes - bull markets, bear markets, high volatility periods - from observed price or return sequences. And in sports analytics, HMMs have been applied to model latent performance states, such as whether a basketball player is "hot" or "cold" during a game based on a sequence of their shot outcomes. This last application, sports analytics, is what we explore in the rest of this tutorial.

---

# 2. Model Specification

## Likelihood

## Priors

## Hierarchical Structure

## Key Modeling Assumptions

---

# 3. Applied Example

## Data Description and Context

We apply a Bayesian Hidden Markov Model to pitch-level velocity data for **Gerrit Cole**, a right-handed starting pitcher for the New York Yankees. Cole was selected for his high four-seam fastball usage rate and consistent full seasons of starts (30+ per year), giving us roughly 1,000–2,000 four-seam fastballs per season to work with.

**Emission metric:** We model a single observable: `release_speed` (pitch velocity in mph) for four-seam fastballs only. Restricting to one pitch type is essential, since mixing pitch types would cause the HMM to discover pitch type as the hidden state rather than performance condition. Velocity is the most straightforward and interpretable signal of a pitcher's condition. Elevated velocity indicates peak effort, while declining velocity often signals fatigue. 

**Hidden states:** The model posits two latent states, *Baseline* and *Hot*, representing the pitcher's true performance mode at each pitch. The state is unobserved; we only see the noisy velocity measurements that each state generates.

**Game structure:** Pitches are grouped by game start, and each game is treated as an **independent sequence**. The hidden state resets at the beginning of each game. This reflects the reality that games are separated by 4–5 days of rest, so carry-over fatigue between games is not modeled.

**Data source:** Pitch-level data come from MLB Statcast via the `baseballr` package, covering the 2019–2024 seasons. The raw data are fetched week-by-week and cached locally to avoid repeated API calls.

### Load Data

We pull pitch-level Statcast data for Gerrit Cole across the 2019–2024 MLB seasons. Data are retrieved in weekly chunks from Baseball Savant and saved locally so subsequent runs read from the cache rather than re-querying the API. All pitch types are retained at this stage; filtering to four-seam fastballs happens in the cleaning step. 

```{r, message=FALSE, warning=FALSE}
library(baseballr)
library(tidyverse)

pitcher_ids <- c(
    "Gerrit Cole" = 543037,
    "Kevin Gausman" = 592332,
    "Logan Webb" = 657277,
    "Zack Wheeler" = 554430,
    "Sandy Alcantara" = 645261
)

# MLB season boundaries (2020 was COVID-shortened)
seasons <- list(
    list(start = "2019-03-28", end = "2019-09-29"),
    list(start = "2020-07-23", end = "2020-09-27"),
    list(start = "2021-04-01", end = "2021-10-03"),
    list(start = "2022-04-07", end = "2022-10-05"),
    list(start = "2023-03-30", end = "2023-10-01"),
    list(start = "2024-04-01", end = "2024-09-29")
)

# Generate week-by-week date ranges across all seasons
date_ranges <- map_dfr(seasons, function(s) {
    start_dates <- seq(as.Date(s$start), as.Date(s$end) - 6, by = "7 days")
    end_dates <- pmin(start_dates + 6, as.Date(s$end))
    tibble(start = start_dates, end = end_dates)
})

fetch_week <- function(start_dt, end_dt, i, n, player_id) {
    message(sprintf("[%d/%d] Pulling: %s to %s", i, n, start_dt, end_dt))

    result <- tryCatch(
        {
            statcast_search(
                start_date  = as.character(start_dt),
                end_date    = as.character(end_dt),
                playerid    = player_id,
                player_type = "pitcher"
            )
        },
        error = function(e) {
            message(sprintf("[%d/%d] Error: %s", i, n, e$message))
            return(NULL)
        }
    )

    if (is.null(result) || nrow(result) == 0) {
        return(NULL)
    }
    Sys.sleep(3)
    return(result)
}

fetch_statcast <- function(player_name, player_id, cache_file) {
    if (file.exists(cache_file)) {
        message("Loading cached data from ", cache_file)
        return(read_csv(cache_file, show_col_types = FALSE))
    }

    message("Scraping Statcast data for ", player_name, " — this will take ~10 minutes")
    n_weeks <- nrow(date_ranges)
    raw_data <- pmap_dfr(
        list(date_ranges$start, date_ranges$end, seq_len(n_weeks), n_weeks, player_id),
        fetch_week
    )

    df <- raw_data %>%
        mutate(season = lubridate::year(as.Date(game_date))) %>%
        select(
            player_name, game_date, season,
            pitch_type,
            inning, at_bat_number, pitch_number,
            release_speed, release_spin_rate,
            pfx_x, pfx_z,
            release_extension,
            balls, strikes
        )

    write_csv(df, cache_file)
    df
}

gerrit_cole <- fetch_statcast("Gerrit Cole", pitcher_ids[["Gerrit Cole"]], "data/gerrit_cole.csv")
```

Below is a sample of the raw data that was collected: 

```{r}
gerrit_cole |>
    select(player_name, game_date, season, pitch_type, release_speed, release_spin_rate) |>
    head() |>
    knitr::kable(col.names = c("Pitcher", "Date", "Season", "Pitch Type", "Velocity (mph)", "Spin Rate (rpm)"))
```

### Clean Data

We filter the raw data to four-seam fastballs for the 2024 season and restructure it into the format expected by the HMM. Within each season, games are numbered sequentially by date. Within each game, pitches are sorted by inning, then at-bat, then position within the at-bat to recover true chronological order, and renumbered 1, 2, 3, ... as a single game-level sequence.

```{r}
metrics_all <- c("release_speed", "release_spin_rate", "pfx_x", "pfx_z", "release_extension")

get_player_data <- function(data, name, pitch_type = NULL, season_filter = NULL, metrics = NULL) {
    if (is.null(metrics)) metrics <- metrics_all

    data %>%
        filter(player_name == name) %>%
        {
            if (!is.null(pitch_type)) filter(., pitch_type %in% !!pitch_type) else .
        } %>%
        {
            if (!is.null(season_filter)) filter(., season %in% season_filter) else .
        } %>%
        group_by(season, game_date) %>%
        arrange(inning, at_bat_number, pitch_number, .by_group = TRUE) %>%
        mutate(pitch_number = row_number()) %>%
        ungroup() %>%
        group_by(season) %>%
        mutate(game_number = dense_rank(game_date)) %>%
        ungroup() %>%
        select(season, player = player_name, game_number, pitch_number, all_of(metrics))
}

filtered_gerrit_cole <- get_player_data(gerrit_cole, "Cole, Gerrit", pitch_type = "FF", season_filter = max(gerrit_cole$season), metrics = "release_speed")
filtered_gerrit_cole |>
    head() |>
    knitr::kable(col.names = c("Season", "Pitcher", "Game", "Pitch", "Velocity (mph)"))
```

\newpage

## Fit Model

### Prior Selection

We need to center the Normal prior on each emission mean somewhere reasonable. As a starting point, we use Cole's historical velocity data from all seasons prior to 2024, split at the 75th percentile: the lower 75% of pitches approximate the Baseline state and the upper 25% approximate the Hot state. The mean of each group becomes the center of the corresponding prior. This is a rough heuristic. A 75/25 split doesn't correspond to any structural assumption about how often Cole is in each state. The intent is simply to give the sampler a sensible starting region rather than an uninformative flat prior. With thousands of 2024 pitches available, the likelihood will dominate and the prior centers matter little in practice.

```{r}
# Use pre-2024 seasons to get empirical prior means for each state
historic_cole <- get_player_data(gerrit_cole, "Cole, Gerrit",
    pitch_type = "FF",
    season_filter = setdiff(unique(gerrit_cole$season), max(gerrit_cole$season)),
    metrics = "release_speed"
)

# Split at the 75th percentile: baseline (lower 75%) vs. hot (upper 25%)
q75 <- quantile(historic_cole$release_speed, 0.75)
baseline_pitches <- historic_cole$release_speed[historic_cole$release_speed <= q75]
hot_pitches <- historic_cole$release_speed[historic_cole$release_speed > q75]

empirical_mu <- c(mean(baseline_pitches), mean(hot_pitches))

tibble(
    State = c("Baseline", "Hot"),
    `Mean (mph)` = round(c(mean(baseline_pitches), mean(hot_pitches)), 2),
    `SD (mph)` = round(c(sd(baseline_pitches), sd(hot_pitches)), 2),
    `N (pitches)` = c(length(baseline_pitches), length(hot_pitches))
) |>
    knitr::kable()

# Round to nearest integer for use as prior means
prior_mu_mean <- c(96, 99)
```

The three sets of model parameters each receive their own prior.

For the **emission means** ($\mu_k$), we use Normal priors centered on the empirical group means above, with a standard deviation of 2 mph. This is wide enough that the posterior is driven by the 2024 data, but anchored close enough to realistic velocity values that the sampler does not waste time exploring implausible regions.

| State    | Prior mean (mph) | Prior SD (mph) |
|----------|-----------------|----------------|
| Baseline | 96              | 2              |
| Hot      | 99              | 2              |

For the **transition matrix rows** ($\nu_k$), we use Dirichlet priors that place more weight on self-transitions than on switching. A concentration parameter of 8 on the diagonal versus 1 on the off-diagonal reflects the prior belief that a pitcher's state tends to persist from one pitch to the next — states are sticky, not erratic.

| Current state | Stay (Dirichlet param) | Switch (Dirichlet param) | Prior P(stay) |
|---------------|------------------------|--------------------------|---------------|
| Baseline      | 8                      | 1                        | ~89%          |
| Hot           | 8                      | 1                        | ~89%          |

For the **initial state distribution** ($\pi$), we use a Dirichlet prior that puts most of its weight on starting in the Baseline state at the beginning of each game. The asymmetry reflects the reasonable assumption that a pitcher is more likely to begin a start in a typical rather than elevated performance mode.

| State    | Dirichlet param | Prior P(start here) |
|----------|-----------------|---------------------|
| Baseline | 6               | ~75%                |
| Hot      | 2               | ~25%                |


\newpage

### Gibbs Sampler

The HMM has three sets of unknown parameters: the emission means $\mu$, the transition matrix $\nu$, and the initial state distribution $\pi$, plus the hidden state sequence itself. Because the full joint posterior over all of these is intractable, we use a **Gibbs sampler**: a Markov chain Monte Carlo algorithm that cycles through each block of unknowns in turn, drawing from its conditional distribution given the current values of everything else. After a burn-in period, the chain's samples approximate the joint posterior.

The hidden state sequences are a fourth unknown, but they are not a parameter in the usual sense. They are latent variables that must be integrated over. Because conditioning on the state sequences makes the parameter updates simple conjugate draws, we include them in the sampler and alternate between updating states and updating parameters.

Each iteration of the sampler performs four steps:

**Step 1: Sample hidden states (FFBS).** Given the current parameter values, we sample a complete hidden state sequence for each game independently using the **Forward-Filter Backward-Sample** algorithm. In the forward pass, we compute $\alpha_{t,k}$, the joint probability of the observed velocities up through pitch $t$ and being in state $k$ at pitch $t$. This recursion sweeps forward through the game, accumulating evidence. In the backward pass, we draw states from right to left: the last pitch's state is sampled from the normalized final forward probabilities, and each earlier state is sampled conditional on the state drawn immediately after it. This produces an exact draw from $p(\mathbf{z} \mid \mathbf{y}, \mu, \nu, \pi)$.

Because games are independent sequences, hidden states reset at the start of each game, with no carry-over between starts. FFBS is run separately for each of the $G$ games per iteration.

**Step 2: Sample transition probabilities ($\nu$).** Given the sampled state sequences, we count how many times each state-to-state transition occurs across all games. These counts, added to the Dirichlet prior parameters, form the posterior hyperparameters for each row of $\nu$. Each row is then drawn independently from its Dirichlet full conditional.

**Step 3: Sample initial state distribution ($\pi$).** We count how many games begin in each state (i.e., look at the first sampled state of each game). Adding these counts to the Dirichlet prior on $\pi$ gives the posterior hyperparameters, from which we draw a new $\pi$.

**Step 4: Sample emission means ($\mu_k$).** Given the state assignments for every pitch, we pool all pitches assigned to state $k$ across all games and compute their sample mean and count. Combined with the Normal prior, this gives a Normal full conditional for $\mu_k$, from which we draw directly. This step is repeated for each state.

We also apply a **label-switching correction** after each iteration. Because the states $\{1, 2\}$ are exchangeable under the likelihood, the sampler can freely permute their labels between iterations. Without correction, samples from different chains would not be comparable. We fix this by enforcing a canonical ordering: after each draw, we sort the states by their emission means ($\mu_1 < \mu_2$), relabeling all state assignments, both rows and columns of $\nu$, and $\pi$ accordingly.

To assess convergence and reduce dependence on starting values, we run **four independent chains** from different initializations and discard the first half of each chain as burn-in. We reshape the pitch-level data into $Y$, a list of length $G$ where $Y_g$ is the ordered vector of four-seam fastball velocities for game $g$. We also fix $\sigma$ to the empirical standard deviation of all 2024 velocities, a shared, known variance assumed by the model.

```{r}
# Reshape cleaned data into list-of-games format expected by the HMM
Y <- split(filtered_gerrit_cole$release_speed, filtered_gerrit_cole$game_number)
Y <- lapply(Y, unname)
sigma <- sd(unlist(Y))

run_gibbs_chain <- function(Y, sigma, prior_nu, prior_pi, prior_mu_mean, prior_mu_var, init_nu, init_pi, init_mu, numsamp) {
    K <- length(init_mu)
    G <- length(Y)
    Tg <- sapply(Y, length)
    total_pitches <- sum(Tg)

    # Storage matrices
    mu.samp <- matrix(NA, nrow = numsamp, ncol = K)
    nu.samp <- matrix(NA, nrow = numsamp, ncol = K * K)
    pi.samp <- matrix(NA, nrow = numsamp, ncol = K)
    X.samp <- matrix(NA, nrow = numsamp, ncol = total_pitches)

    # Current parameter values
    curnu <- init_nu
    curpi <- init_pi
    curmu <- init_mu
    X <- lapply(Tg, function(t) rep(NA, t))

    for (iter in 1:numsamp) {
        #### Block 1: FFBS — sample hidden states per game ####
        for (g in 1:G) {
            alpha <- matrix(NA, nrow = Tg[g], ncol = K)

            # Forward pass
            for (i in 1:K) {
                alpha[1, i] <- curpi[i] * dnorm(Y[[g]][1], curmu[i], sigma)
            }
            if (Tg[g] > 1) {
                for (t in 2:Tg[g]) {
                    for (i in 1:K) {
                        alpha[t, i] <- 0
                        for (j in 1:K) {
                            alpha[t, i] <- alpha[t, i] +
                                alpha[t - 1, j] * curnu[j, i] * dnorm(Y[[g]][t], curmu[i], sigma)
                        }
                    }
                }
            }

            # Backward sampling
            prob_vec <- alpha[Tg[g], ] / sum(alpha[Tg[g], ])
            X[[g]][Tg[g]] <- sample(1:K, size = 1, prob = prob_vec)
            if (Tg[g] > 1) {
                for (t in (Tg[g] - 1):1) {
                    for (i in 1:K) {
                        prob_vec[i] <- alpha[t, i] * curnu[i, X[[g]][t + 1]]
                    }
                    prob_vec <- prob_vec / sum(prob_vec)
                    X[[g]][t] <- sample(1:K, size = 1, prob = prob_vec)
                }
            }
        }

        #### Block 2: Sample transition matrix ####
        Nmat <- matrix(0, nrow = K, ncol = K)
        for (g in 1:G) {
            for (t in 2:Tg[g]) {
                j <- X[[g]][t - 1]
                k <- X[[g]][t]
                Nmat[j, k] <- Nmat[j, k] + 1
            }
        }
        for (j in 1:K) {
            for (k in 1:K) {
                curnu[j, k] <- rgamma(1, shape = Nmat[j, k] + prior_nu[j, k], rate = 1)
            }
            curnu[j, ] <- curnu[j, ] / sum(curnu[j, ])
        }

        #### Block 3: Sample initial distribution ####
        init_count <- rep(0, K)
        for (g in 1:G) {
            init_count[X[[g]][1]] <- init_count[X[[g]][1]] + 1
        }
        for (j in 1:K) {
            curpi[j] <- rgamma(1, shape = init_count[j] + prior_pi[j], rate = 1)
        }
        curpi <- curpi / sum(curpi)

        #### Block 4: Sample emission means ####
        for (j in 1:K) {
            obs_j <- unlist(lapply(1:G, function(g) Y[[g]][X[[g]] == j]))
            n_j <- length(obs_j)
            if (n_j > 0) {
                Y_bar_j <- mean(obs_j)
                post_prec <- 1 / prior_mu_var + n_j / sigma^2
                post_mean <- (prior_mu_mean[j] / prior_mu_var + n_j * Y_bar_j / sigma^2) / post_prec
                curmu[j] <- rnorm(1, mean = post_mean, sd = sqrt(1 / post_prec))
            } else {
                curmu[j] <- rnorm(1, mean = prior_mu_mean[j], sd = sqrt(prior_mu_var))
            }
        }

        #### Store samples ####
        mu.samp[iter, ] <- curmu
        nu.samp[iter, ] <- as.vector(curnu)
        pi.samp[iter, ] <- curpi
        X.samp[iter, ] <- unlist(X)
    }

    list(mu = mu.samp, nu = nu.samp, pi = pi.samp, X = X.samp, game_lengths = Tg)
}

fix_label_switching <- function(mu_mat, nu_mat, pi_mat, X_mat) {
    n_iter <- nrow(mu_mat)
    for (i in 1:n_iter) {
        if (mu_mat[i, 1] > mu_mat[i, 2]) {
            # Swap mu
            temp_mu <- mu_mat[i, 1]
            mu_mat[i, 1] <- mu_mat[i, 2]
            mu_mat[i, 2] <- temp_mu

            # Swap pi
            temp_pi <- pi_mat[i, 1]
            pi_mat[i, 1] <- pi_mat[i, 2]
            pi_mat[i, 2] <- temp_pi

            # Swap nu (column-major vectorization: nu11, nu21, nu12, nu22)
            temp_nu11 <- nu_mat[i, 1]
            temp_nu21 <- nu_mat[i, 2]
            temp_nu12 <- nu_mat[i, 3]
            temp_nu22 <- nu_mat[i, 4]
            nu_mat[i, 1] <- temp_nu22
            nu_mat[i, 4] <- temp_nu11
            nu_mat[i, 2] <- temp_nu12
            nu_mat[i, 3] <- temp_nu21

            # Relabel X
            current_X <- X_mat[i, ]
            X_mat[i, current_X == 1] <- 2
            X_mat[i, current_X == 2] <- 1
        }
    }
    list(mu = mu_mat, nu = nu_mat, pi = pi_mat, X = X_mat)
}
```


```{r, message=FALSE}
# Priors
K <- 2
prior_nu <- rbind(c(8, 1), c(1, 8))
prior_pi <- c(6, 2)
prior_mu_var <- 4 # SD = 2 mph
numsamp <- 10000
y_mean <- mean(unlist(Y))

cache_file <- "data/gerrit_cole_chains.rds"

if (file.exists(cache_file)) {
    message("Loading cached chains from ", cache_file)
    chains <- readRDS(cache_file)
    chain1 <- chains$chain1
    chain2 <- chains$chain2
    chain3 <- chains$chain3
    chain4 <- chains$chain4
} else {
    # Chain 1: near-prior starting values
    chain1 <- run_gibbs_chain(
        Y, sigma, prior_nu, prior_pi, prior_mu_mean, prior_mu_var,
        init_nu = rbind(c(0.75, 0.25), c(0.25, 0.75)),
        init_pi = c(0.5, 0.5),
        init_mu = c(y_mean - 1, y_mean + 1),
        numsamp = numsamp
    )

    # Chain 2: dispersed starting values
    chain2 <- run_gibbs_chain(
        Y, sigma, prior_nu, prior_pi, prior_mu_mean, prior_mu_var,
        init_nu = rbind(c(0.5, 0.5), c(0.5, 0.5)),
        init_pi = c(0.75, 0.25),
        init_mu = c(y_mean - 2, y_mean + 2),
        numsamp = numsamp
    )

    # Chain 3: strongly dispersed starting values
    chain3 <- run_gibbs_chain(
        Y, sigma, prior_nu, prior_pi, prior_mu_mean, prior_mu_var,
        init_nu = rbind(c(0.9, 0.1), c(0.1, 0.9)),
        init_pi = c(0.25, 0.75),
        init_mu = c(y_mean - 3, y_mean + 3),
        numsamp = numsamp
    )

    # Chain 4: near-uniform starting values
    chain4 <- run_gibbs_chain(
        Y, sigma, prior_nu, prior_pi, prior_mu_mean, prior_mu_var,
        init_nu = rbind(c(0.6, 0.4), c(0.4, 0.6)),
        init_pi = c(0.5, 0.5),
        init_mu = c(y_mean - 0.5, y_mean + 0.5),
        numsamp = numsamp
    )

    saveRDS(list(chain1 = chain1, chain2 = chain2, chain3 = chain3, chain4 = chain4), cache_file)
    message("Chains saved to ", cache_file)
}

# Fix label switching in each chain
fixed1 <- fix_label_switching(chain1$mu, chain1$nu, chain1$pi, chain1$X)
fixed2 <- fix_label_switching(chain2$mu, chain2$nu, chain2$pi, chain2$X)
fixed3 <- fix_label_switching(chain3$mu, chain3$nu, chain3$pi, chain3$X)
fixed4 <- fix_label_switching(chain4$mu, chain4$nu, chain4$pi, chain4$X)
```

\newpage

## MCMC Diagnostics

We ran four independent Gibbs chains from different starting values to check convergence. Trace plots show the sampled value of each parameter at every iteration. Four well-mixed chains that overlap and wander around the same region, rather than drifting apart or getting stuck, are the primary visual signal that the sampler has converged.

We monitor all sampled scalar parameters across three figures: emission means ($\mu_1$, $\mu_2$), all four transition probabilities ($\nu_{11}$, $\nu_{12}$, $\nu_{21}$, $\nu_{22}$), and the initial state distribution ($\pi_1$, $\pi_2$).

### Initial Samples

These are the raw trace plots from the Gibbs sampler: every draw from all 10,000 iterations, with no post-processing applied. Consecutive samples from a Gibbs chain are not independent: each draw depends on the one before it, so the chain can exhibit autocorrelation, where successive values move slowly and the effective number of independent samples is much smaller than the raw count.

```{r trace-plot-fns, include=FALSE}
fixed_chains <- list(
    "Chain 1" = fixed1, "Chain 2" = fixed2,
    "Chain 3" = fixed3, "Chain 4" = fixed4
)

plot_trace_mu <- function(chains, numsamp) {
    mu_long <- bind_rows(
        imap(chains, ~ tibble(
            iteration = 1:numsamp, chain = .y,
            `mu1 (Baseline)` = .x$mu[, 1], `mu2 (Hot)` = .x$mu[, 2]
        ))
    ) |>
        pivot_longer(cols = -c(iteration, chain), names_to = "parameter", values_to = "value") |>
        mutate(parameter = factor(parameter, levels = c("mu1 (Baseline)", "mu2 (Hot)")))

    ggplot(mu_long, aes(x = iteration, y = value, color = chain)) +
        geom_line(linewidth = 0.3, alpha = 0.9) +
        facet_wrap(~parameter, ncol = 2, scales = "fixed") +
        coord_cartesian(ylim = c(94, 98)) +
        labs(title = "Emission Means", x = "Iteration", y = "mph", color = NULL) +
        theme_bw() +
        theme(legend.position = "right", strip.text = element_text(face = "bold"))
}

plot_trace_nu <- function(chains, numsamp) {
    nu_long <- bind_rows(
        imap(chains, ~ tibble(
            iteration = 1:numsamp, chain = .y,
            `nu11 (Baseline->Baseline)` = .x$nu[, 1],
            `nu12 (Baseline->Hot)` = .x$nu[, 3],
            `nu21 (Hot->Baseline)` = .x$nu[, 2],
            `nu22 (Hot->Hot)` = .x$nu[, 4]
        ))
    ) |>
        pivot_longer(cols = -c(iteration, chain), names_to = "parameter", values_to = "value") |>
        mutate(parameter = factor(parameter, levels = c(
            "nu11 (Baseline->Baseline)", "nu12 (Baseline->Hot)",
            "nu21 (Hot->Baseline)",      "nu22 (Hot->Hot)"
        )))

    ggplot(nu_long, aes(x = iteration, y = value, color = chain)) +
        geom_line(linewidth = 0.3, alpha = 0.9) +
        facet_wrap(~parameter, ncol = 2, scales = "fixed") +
        coord_cartesian(ylim = c(0, 1)) +
        labs(title = "Transition Probabilities", x = "Iteration", y = "Probability", color = NULL) +
        theme_bw() +
        theme(legend.position = "right", strip.text = element_text(face = "bold"))
}

plot_trace_pi <- function(chains, numsamp) {
    pi_long <- bind_rows(
        imap(chains, ~ tibble(
            iteration = 1:numsamp, chain = .y,
            `pi1 (Baseline)` = .x$pi[, 1], `pi2 (Hot)` = .x$pi[, 2]
        ))
    ) |>
        pivot_longer(cols = -c(iteration, chain), names_to = "parameter", values_to = "value") |>
        mutate(parameter = factor(parameter, levels = c("pi1 (Baseline)", "pi2 (Hot)")))

    ggplot(pi_long, aes(x = iteration, y = value, color = chain)) +
        geom_line(linewidth = 0.3, alpha = 0.9) +
        facet_wrap(~parameter, ncol = 2, scales = "fixed") +
        coord_cartesian(ylim = c(0, 1)) +
        labs(title = "Initial State Distribution", x = "Iteration", y = "Probability", color = NULL) +
        theme_bw() +
        theme(legend.position = "right", strip.text = element_text(face = "bold"))
}

plot_acf_mu <- function(chains) {
    ci <- qnorm(0.975) / sqrt(nrow(chains[[1]]$mu))
    ch <- chains[[1]]
    acf_data <- bind_rows(
        {
            a <- acf(ch$mu[, 1], plot = FALSE)
            tibble(lag = as.numeric(a$lag), acf = as.numeric(a$acf), parameter = "mu1 (Baseline)")
        },
        {
            a <- acf(ch$mu[, 2], plot = FALSE)
            tibble(lag = as.numeric(a$lag), acf = as.numeric(a$acf), parameter = "mu2 (Hot)")
        }
    ) |>
        mutate(parameter = factor(parameter, levels = c("mu1 (Baseline)", "mu2 (Hot)")))

    ggplot(acf_data, aes(x = lag, y = acf)) +
        geom_hline(yintercept = 0) +
        geom_segment(aes(xend = lag, yend = 0)) +
        geom_hline(yintercept = c(-ci, ci), linetype = "dashed", color = "steelblue") +
        facet_wrap(~parameter, ncol = 2) +
        labs(title = "ACF: Emission Means", x = "Lag", y = "ACF") +
        theme_bw() +
        theme(strip.text = element_text(face = "bold"))
}

plot_acf_nu <- function(chains) {
    ci <- qnorm(0.975) / sqrt(nrow(chains[[1]]$nu))
    ch <- chains[[1]]
    acf_data <- bind_rows(
        {
            a <- acf(ch$nu[, 1], plot = FALSE)
            tibble(lag = as.numeric(a$lag), acf = as.numeric(a$acf), parameter = "nu11 (Baseline->Baseline)")
        },
        {
            a <- acf(ch$nu[, 3], plot = FALSE)
            tibble(lag = as.numeric(a$lag), acf = as.numeric(a$acf), parameter = "nu12 (Baseline->Hot)")
        },
        {
            a <- acf(ch$nu[, 2], plot = FALSE)
            tibble(lag = as.numeric(a$lag), acf = as.numeric(a$acf), parameter = "nu21 (Hot->Baseline)")
        },
        {
            a <- acf(ch$nu[, 4], plot = FALSE)
            tibble(lag = as.numeric(a$lag), acf = as.numeric(a$acf), parameter = "nu22 (Hot->Hot)")
        }
    ) |>
        mutate(parameter = factor(parameter, levels = c(
            "nu11 (Baseline->Baseline)", "nu12 (Baseline->Hot)",
            "nu21 (Hot->Baseline)",      "nu22 (Hot->Hot)"
        )))

    ggplot(acf_data, aes(x = lag, y = acf)) +
        geom_hline(yintercept = 0) +
        geom_segment(aes(xend = lag, yend = 0)) +
        geom_hline(yintercept = c(-ci, ci), linetype = "dashed", color = "steelblue") +
        facet_wrap(~parameter, ncol = 2) +
        labs(title = "ACF: Transition Probabilities", x = "Lag", y = "ACF") +
        theme_bw() +
        theme(strip.text = element_text(face = "bold"))
}

plot_acf_pi <- function(chains) {
    ci <- qnorm(0.975) / sqrt(nrow(chains[[1]]$pi))
    ch <- chains[[1]]
    acf_data <- bind_rows(
        {
            a <- acf(ch$pi[, 1], plot = FALSE)
            tibble(lag = as.numeric(a$lag), acf = as.numeric(a$acf), parameter = "pi1 (Baseline)")
        },
        {
            a <- acf(ch$pi[, 2], plot = FALSE)
            tibble(lag = as.numeric(a$lag), acf = as.numeric(a$acf), parameter = "pi2 (Hot)")
        }
    ) |>
        mutate(parameter = factor(parameter, levels = c("pi1 (Baseline)", "pi2 (Hot)")))

    ggplot(acf_data, aes(x = lag, y = acf)) +
        geom_hline(yintercept = 0) +
        geom_segment(aes(xend = lag, yend = 0)) +
        geom_hline(yintercept = c(-ci, ci), linetype = "dashed", color = "steelblue") +
        facet_wrap(~parameter, ncol = 2) +
        labs(title = "ACF: Initial State Distribution", x = "Lag", y = "ACF") +
        theme_bw() +
        theme(strip.text = element_text(face = "bold"))
}
```

```{r trace-mu, fig.height=3}
plot_trace_mu(fixed_chains, numsamp)
```

```{r trace-nu, fig.height=5}
plot_trace_nu(fixed_chains, numsamp)
```

```{r trace-pi, fig.height=3}
plot_trace_pi(fixed_chains, numsamp)
```

\newpage

### Thinned Samples

To diagnose autocorrelation, we compute the **autocorrelation function (ACF)** for each parameter. The ACF measures the correlation between a chain's samples at lag $k$, that is, how similar a draw is to the draw $k$ steps before it. High autocorrelation at moderate lags (say, lags 1–10) indicates that the chain is moving slowly through the posterior and that many consecutive samples carry nearly the same information. The lag at which the ACF first drops close to zero tells us the minimum spacing needed for approximately independent draws.

```{r acf-mu, fig.height=3}
plot_acf_mu(fixed_chains)
```

```{r acf-nu, fig.height=5}
plot_acf_nu(fixed_chains)
```

```{r acf-pi, fig.height=3}
plot_acf_pi(fixed_chains)
```

Based on the ACF, we thin the chain by keeping only every $k$-th sample, where $k$ is chosen to be at or beyond the lag where autocorrelation becomes negligible. Thinning reduces the effective sample size but produces a set of draws that are much closer to independent, which makes posterior summaries and credible intervals more reliable. The trace plots below show the sampled values after thinning.

The ACF plots above show that autocorrelation across all parameters decays fully by roughly lag 8–10. We discard the first 1,000 iterations as burn-in to ensure the starting values have no influence on inference, then thin by keeping every 10th sample. This leaves 900 draws per chain (3,600 combined), which is enough for stable posterior summaries.

```{r thinning}
burn_in <- 1000
thin_k <- 10
postburn <- (burn_in + 1):numsamp
thin_idx <- postburn[seq(1, length(postburn), by = thin_k)]

thinned_chains <- lapply(fixed_chains, function(ch) {
    list(
        mu = ch$mu[thin_idx, ],
        nu = ch$nu[thin_idx, ],
        pi = ch$pi[thin_idx, ],
        # ch$X is a (numsamp x total_pitches) matrix; drop = FALSE keeps it
        # 2D when thin_idx happens to be length 1 and prevents column-major
        # linear indexing from silently returning the wrong elements.
        X  = ch$X[thin_idx, , drop = FALSE]
    )
})
numsamp_thinned <- length(thin_idx)
```

```{r thinned-trace-mu, fig.height=3}
plot_trace_mu(thinned_chains, numsamp_thinned)
```

```{r thinned-trace-nu, fig.height=5}
plot_trace_nu(thinned_chains, numsamp_thinned)
```

```{r thinned-trace-pi, fig.height=3}
plot_trace_pi(thinned_chains, numsamp_thinned)
```

## Posterior Summaries

```{r posterior-summaries}
# Pool thinned draws across all four chains
mu_pool <- do.call(rbind, lapply(thinned_chains, `[[`, "mu"))
nu_pool <- do.call(rbind, lapply(thinned_chains, `[[`, "nu"))
pi_pool <- do.call(rbind, lapply(thinned_chains, `[[`, "pi"))

summarise_param <- function(draws, name) {
    tibble(
        Parameter = name,
        Mean      = mean(draws),
        SD        = sd(draws),
        `10%`     = quantile(draws, 0.10),
        `90%`     = quantile(draws, 0.90)
    )
}

posterior_table <- bind_rows(
    summarise_param(mu_pool[, 1], "mu1 (Baseline mean, mph)"),
    summarise_param(mu_pool[, 2], "mu2 (Hot mean, mph)"),
    summarise_param(nu_pool[, 1], "nu11 (Baseline -> Baseline)"),
    summarise_param(nu_pool[, 3], "nu12 (Baseline -> Hot)"),
    summarise_param(nu_pool[, 2], "nu21 (Hot -> Baseline)"),
    summarise_param(nu_pool[, 4], "nu22 (Hot -> Hot)"),
    summarise_param(pi_pool[, 1], "pi1 (Pr. start Baseline)"),
    summarise_param(pi_pool[, 2], "pi2 (Pr. start Hot)")
)

knitr::kable(
    posterior_table,
    digits  = 3,
    caption = "Posterior summaries from 3,600 thinned draws (4 chains × 900 samples). CI bounds are the 10th and 90th percentiles."
)
```

\newpage

## Posterior Predictive Checks

## Visualizations

The plot below pairs observed fastball velocity (top row) with the posterior $P(\text{Hot})$ for each pitch (bottom row), one column per game. The three games are chosen to span the within-game $\text{sd}(P(\text{Hot}))$ distribution (minimum, median, and maximum) so reading left-to-right tours the full range of HMM behavior, from starts where the model is entirely confident to starts where it is actively switching states. Reading each column *vertically* is the central diagnostic: a pitch near $\hat\mu_{\text{Hot}}$ should line up with a point near 1 directly below it, and a pitch near $\hat\mu_{\text{Baseline}}$ with a point near 0.

**How this lines up with reality:** Cole's 2024 was cut in half by an elbow injury that sidelined him through mid-June, which is why "game 2" in this plot is already June 25 and "game 9" lands on August 10. The two starts bracket his comeback arc. Game 2 was his second outing back, a Subway Series loss in which he struggled and next-day coverage ran with headlines like *"Gerrit Cole's Subway Series clobbering comes with stark velocity concerns"* (Joyce, *NY Post*, 2024-06-25). Game 9 was the opposite: six scoreless innings and a season-high ten strikeouts against Texas, widely reported as a return to ace form. The HMM never saw a box score or a headline, it only saw release speeds, but the pitch-level posterior on these two starts already encodes what the writers were saying at the time, which is a useful sanity check that the latent "Hot" state is picking up something real.

```{r viz-state-velocity-stack, fig.height=4}
# Pool thinned X across chains and compute per-pitch P(Hot). This block is
# self-contained for the visualization section; it can be hoisted into a
# dedicated State Decoding section later without changing anything here.
X_pool <- do.call(rbind, lapply(thinned_chains, `[[`, "X"))
stopifnot(ncol(X_pool) == sum(sapply(Y, length)))
p_hot_flat <- colMeans(X_pool == 2)

posterior_states <- filtered_gerrit_cole |>
    arrange(game_number, pitch_number) |>
    mutate(p_hot = p_hot_flat)

mu_baseline_hat <- mean(mu_pool[, 1])
mu_hot_hat <- mean(mu_pool[, 2])

# Pick 3 games that span the within-game sd(p_hot) distribution: the
# minimum, median, and maximum sd among games with >= 30 pitches. Reading
# the facet columns left-to-right gives flat -> mixed -> dynamic, which
# shows the full range of HMM behavior across the season (and doubles as
# a sanity check: the min-sd game should have a nearly flat P(Hot) panel).
games_to_show <- posterior_states |>
    group_by(game_number) |>
    summarise(n_pitches = n(), sd_p = sd(p_hot), .groups = "drop") |>
    filter(n_pitches >= 30) |>
    arrange(sd_p) |>
    slice(c(1, ceiling(n() / 2), n())) |>
    pull(game_number)

# Long format: two rows per pitch (velocity + P(Hot)). facet_grid with
# scales = "free_y" gives each metric row its own units — the native
# ggplot equivalent of par(mfrow).
plot_df <- posterior_states |>
    filter(game_number %in% games_to_show) |>
    select(game_number, pitch_number, release_speed, p_hot) |>
    # Factor levels in games_to_show order so the facet columns read
    # left-to-right as flat -> mixed -> dynamic instead of by game number.
    mutate(game_number = factor(game_number, levels = games_to_show)) |>
    pivot_longer(
        cols      = c(release_speed, p_hot),
        names_to  = "metric",
        values_to = "value"
    ) |>
    mutate(metric = factor(
        metric,
        levels = c("release_speed", "p_hot"),
        labels = c("Velocity (mph)", "P(Hot)")
    ))

# Reference lines: mu_Baseline and mu_Hot on the velocity panel, 0.5 decision
# threshold on the P(Hot) panel. Tagged with `metric` so facet_grid only
# draws each line in the correct row.
ref_lines <- tibble(
    metric = factor(
        c("Velocity (mph)", "Velocity (mph)", "P(Hot)"),
        levels = levels(plot_df$metric)
    ),
    y = c(mu_baseline_hat, mu_hot_hat, 0.5)
)

ggplot(plot_df, aes(x = pitch_number, y = value)) +
    geom_hline(
        data = ref_lines,
        aes(yintercept = y),
        linetype = "dashed", color = "grey40", linewidth = 0.4
    ) +
    geom_point(aes(color = metric), size = 1.4, show.legend = FALSE) +
    scale_color_manual(values = c(
        "Velocity (mph)" = "black",
        "P(Hot)" = "#4A90D9"
    )) +
    facet_grid(
        metric ~ game_number,
        scales = "free_y",
        switch = "y",
        labeller = labeller(game_number = function(g) paste("Game", g))
    ) +
    labs(x = "Pitch number within game", y = NULL) +
    theme_bw() +
    theme(
        strip.placement = "outside",
        strip.background = element_rect(fill = "grey95", color = NA),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.8, "lines")
    )
```

\newpage

The figure below asks a simpler question: *on average, how hot is Cole as a function of pitch index within a start?* For each pitch position $t$, and separately in every Gibbs draw, we compute the fraction of games (among those that reached pitch $t$) that were in the Hot state at that pitch. The posterior mean of that quantity gives the central line; the 2.5/97.5 quantiles across draws give a 95% credible interval on the across-game mean.

The curve tells a clean warm-up story. The first few pitches sit deep in the Baseline state, the estimate climbs monotonically and crosses 0.5 early in the outing, and from roughly the middle of a start onward it plateaus above 0.5 for the rest of the game. This is the familiar pattern of a pitcher getting loose in the first inning, now showing up in the pitch-level posterior. The credible interval is tightest through the middle of a start, where dozens of games contribute, and widens at both ends where fewer games support the estimate.

```{r viz-mean-p-hot, fig.height=3.2}
# For each pitch position t within a game, compute the across-game mean of
# P(Hot) *separately in every Gibbs draw*, then summarize that distribution.
# The posterior mean gives the point estimate; the 2.5/97.5 quantiles give a
# 95% credible interval on "fraction of games that were Hot at pitch t."

game_lengths <- sapply(Y, length)
game_ends <- cumsum(game_lengths)
game_starts <- c(1, head(game_ends, -1) + 1)
G <- length(Y)
n_draws <- nrow(X_pool)
max_len <- max(game_lengths)

# Accumulate, for each (draw, pitch position), the number of games in state
# Hot and the number of games that reached that pitch. hot_by_draw[s, t] is
# then the across-game mean P(Hot) at pitch t in draw s.
hot_sum <- matrix(0, nrow = n_draws, ncol = max_len)
n_games_t <- integer(max_len)
for (g in seq_len(G)) {
    Lg <- game_lengths[g]
    Xg <- X_pool[, game_starts[g]:game_ends[g], drop = FALSE]
    hot_sum[, seq_len(Lg)] <- hot_sum[, seq_len(Lg)] + (Xg == 2)
    n_games_t[seq_len(Lg)] <- n_games_t[seq_len(Lg)] + 1L
}
hot_by_draw <- sweep(hot_sum, 2, n_games_t, `/`)

# Posterior mean and 95% credible interval across Gibbs draws, then drop the
# deep-in-game tail where only a handful of long starts contribute.
ci <- apply(hot_by_draw, 2, quantile, probs = c(0.025, 0.975))
mean_p_hot_by_t <- tibble(
    pitch_number = seq_len(max_len),
    n_games = n_games_t,
    mean_p_hot = colMeans(hot_by_draw),
    p_lo = ci[1, ],
    p_hi = ci[2, ]
) |>
    filter(n_games >= 10)

ggplot(mean_p_hot_by_t, aes(x = pitch_number, y = mean_p_hot)) +
    geom_hline(
        yintercept = 0.5,
        linetype = "dashed", color = "grey40", linewidth = 0.4
    ) +
    geom_ribbon(
        aes(ymin = p_lo, ymax = p_hi),
        fill = "#4A90D9", alpha = 0.25
    ) +
    geom_line(color = "#1F4E79", linewidth = 0.6) +
    geom_point(color = "#1F4E79", size = 1.3) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        x = "Pitch number within game",
        y = "Mean posterior P(Hot) with 95% CI"
    ) +
    theme_bw()
```

---

# 4. Interpretation

## What Did We Learn?

## Limitations

## When Should We Use This Model?

## Sensitivity to Modeling Assumptions
