---
title: "Bayesian Hidden Markov Models to Infer Latent Pitcher States"
author: "Garrett Lee, Trey Elder, Isaac Herman, Injun Lyo"
date: ""
output:
  pdf_document:
    extra_dependencies: ["float"]
---

# 1. Motivation

## What Problem Does This Method Solve?

## Why Can't Standard Regression Handle It?

## Where Does It Show Up in Practice?

---

# 2. Model Specification

## Likelihood

## Priors

## Hierarchical Structure

## Key Modeling Assumptions

---

# 3. Applied Example

## Data Description and Context

## Load Data

We pull pitch-level Statcast data across the 2019–2024 MLB seasons using `fetch_statcast()`, which wraps a week-by-week API loop via `baseballr` and caches results to a local CSV. Querying by individual player ID rather than pulling all pitchers keeps each API call small. Pitch-type filtering is deferred to the Clean Data step so the raw cache stays flexible. Currently configured for Gerrit Cole.

```{r, message=FALSE, warning=FALSE}
library(baseballr)
library(tidyverse)

pitcher_ids <- c(
    "Gerrit Cole" = 543037,
    "Kevin Gausman" = 592332,
    "Logan Webb" = 657277,
    "Zack Wheeler" = 554430,
    "Sandy Alcantara" = 645261
)

# MLB season boundaries (2020 was COVID-shortened)
seasons <- list(
    list(start = "2019-03-28", end = "2019-09-29"),
    list(start = "2020-07-23", end = "2020-09-27"),
    list(start = "2021-04-01", end = "2021-10-03"),
    list(start = "2022-04-07", end = "2022-10-05"),
    list(start = "2023-03-30", end = "2023-10-01"),
    list(start = "2024-04-01", end = "2024-09-29")
)

# Generate week-by-week date ranges across all seasons
date_ranges <- map_dfr(seasons, function(s) {
    start_dates <- seq(as.Date(s$start), as.Date(s$end) - 6, by = "7 days")
    end_dates <- pmin(start_dates + 6, as.Date(s$end))
    tibble(start = start_dates, end = end_dates)
})

fetch_week <- function(start_dt, end_dt, i, n, player_id) {
    message(sprintf("[%d/%d] Pulling: %s to %s", i, n, start_dt, end_dt))

    result <- tryCatch(
        {
            statcast_search(
                start_date  = as.character(start_dt),
                end_date    = as.character(end_dt),
                playerid    = player_id,
                player_type = "pitcher"
            )
        },
        error = function(e) {
            message(sprintf("[%d/%d] Error: %s", i, n, e$message))
            return(NULL)
        }
    )

    if (is.null(result) || nrow(result) == 0) {
        return(NULL)
    }
    Sys.sleep(3)
    return(result)
}

fetch_statcast <- function(player_name, player_id, cache_file) {
    if (file.exists(cache_file)) {
        message("Loading cached data from ", cache_file)
        return(read_csv(cache_file, show_col_types = FALSE))
    }

    message("Scraping Statcast data for ", player_name, " — this will take ~10 minutes")
    n_weeks <- nrow(date_ranges)
    raw_data <- pmap_dfr(
        list(date_ranges$start, date_ranges$end, seq_len(n_weeks), n_weeks, player_id),
        fetch_week
    )

    df <- raw_data %>%
        mutate(season = lubridate::year(as.Date(game_date))) %>%
        select(
            player_name, game_date, season,
            pitch_type,
            inning, at_bat_number, pitch_number,
            release_speed, release_spin_rate,
            pfx_x, pfx_z,
            release_extension,
            balls, strikes
        )

    write_csv(df, cache_file)
    df
}

gerrit_cole <- fetch_statcast("Gerrit Cole", pitcher_ids[["Gerrit Cole"]], "data/gerrit_cole.csv")
```

## Clean Data

`get_player_data()` filters a raw Statcast dataframe to a single pitcher, optionally subsets by pitch type and season, and restructures the result into the format expected by the HMM. Within each season, games are numbered sequentially by date. Within each game, pitches are sorted by `inning → at_bat_number → pitch_number` (the within-at-bat index) to recover true chronological order, then renumbered 1, 2, 3, ... as a single game-level sequence. The ordering columns are dropped; only the derived `pitch_number` is retained alongside the selected metrics.

```{r}
metrics_all <- c("release_speed", "release_spin_rate", "pfx_x", "pfx_z", "release_extension")

get_player_data <- function(data, name, pitch_type = NULL, season_filter = NULL, metrics = NULL) {
    if (is.null(metrics)) metrics <- metrics_all

    data %>%
        filter(player_name == name) %>%
        {
            if (!is.null(pitch_type)) filter(., pitch_type %in% !!pitch_type) else .
        } %>%
        {
            if (!is.null(season_filter)) filter(., season %in% season_filter) else .
        } %>%
        group_by(season, game_date) %>%
        arrange(inning, at_bat_number, pitch_number, .by_group = TRUE) %>%
        mutate(pitch_number = row_number()) %>%
        ungroup() %>%
        group_by(season) %>%
        mutate(game_number = dense_rank(game_date)) %>%
        ungroup() %>%
        select(season, player = player_name, game_number, pitch_number, all_of(metrics))
}

filtered_gerrit_cole <- get_player_data(gerrit_cole, "Cole, Gerrit", pitch_type = "FF", season_filter = max(df$season), metrics = "release_speed")
head(filtered_gerrit_cole)
```

## Fit Model

## MCMC Diagnostics

## Posterior Summaries

## Posterior Predictive Checks

## Visualizations

---

# 4. Interpretation

## What Did We Learn?

## Limitations

## When Should We Use This Model?

## Sensitivity to Modeling Assumptions
