---
title: "Bayesian Hidden Markov Models to Infer Latent Pitcher States"
author: "Garrett Lee, Trey Elder, Isaac Herman, Injun Lyo"
date: ""
output:
  pdf_document:
    extra_dependencies: ["float"]
---

# 1. Motivation

## What Problem Does This Method Solve?

## Why Can't Standard Regression Handle It?

## Where Does It Show Up in Practice?

---

# 2. Model Specification

## Likelihood

## Priors

## Hierarchical Structure

## Key Modeling Assumptions

---

# 3. Applied Example

## Data Description and Context

## Load Data

We pull pitch-level Statcast data across the 2019–2024 MLB seasons using `fetch_statcast()`, which wraps a week-by-week API loop via `baseballr` and caches results to a local CSV. Querying by individual player ID rather than pulling all pitchers keeps each API call small. Pitch-type filtering is deferred to the Clean Data step so the raw cache stays flexible. Currently configured for Gerrit Cole.

```{r, message=FALSE, warning=FALSE}
library(baseballr)
library(tidyverse)

pitcher_ids <- c(
    "Gerrit Cole" = 543037,
    "Kevin Gausman" = 592332,
    "Logan Webb" = 657277,
    "Zack Wheeler" = 554430,
    "Sandy Alcantara" = 645261
)

# MLB season boundaries (2020 was COVID-shortened)
seasons <- list(
    list(start = "2019-03-28", end = "2019-09-29"),
    list(start = "2020-07-23", end = "2020-09-27"),
    list(start = "2021-04-01", end = "2021-10-03"),
    list(start = "2022-04-07", end = "2022-10-05"),
    list(start = "2023-03-30", end = "2023-10-01"),
    list(start = "2024-04-01", end = "2024-09-29")
)

# Generate week-by-week date ranges across all seasons
date_ranges <- map_dfr(seasons, function(s) {
    start_dates <- seq(as.Date(s$start), as.Date(s$end) - 6, by = "7 days")
    end_dates <- pmin(start_dates + 6, as.Date(s$end))
    tibble(start = start_dates, end = end_dates)
})

fetch_week <- function(start_dt, end_dt, i, n, player_id) {
    message(sprintf("[%d/%d] Pulling: %s to %s", i, n, start_dt, end_dt))

    result <- tryCatch(
        {
            statcast_search(
                start_date  = as.character(start_dt),
                end_date    = as.character(end_dt),
                playerid    = player_id,
                player_type = "pitcher"
            )
        },
        error = function(e) {
            message(sprintf("[%d/%d] Error: %s", i, n, e$message))
            return(NULL)
        }
    )

    if (is.null(result) || nrow(result) == 0) {
        return(NULL)
    }
    Sys.sleep(3)
    return(result)
}

fetch_statcast <- function(player_name, player_id, cache_file) {
    if (file.exists(cache_file)) {
        message("Loading cached data from ", cache_file)
        return(read_csv(cache_file, show_col_types = FALSE))
    }

    message("Scraping Statcast data for ", player_name, " — this will take ~10 minutes")
    n_weeks <- nrow(date_ranges)
    raw_data <- pmap_dfr(
        list(date_ranges$start, date_ranges$end, seq_len(n_weeks), n_weeks, player_id),
        fetch_week
    )

    df <- raw_data %>%
        mutate(season = lubridate::year(as.Date(game_date))) %>%
        select(
            player_name, game_date, season,
            pitch_type,
            inning, at_bat_number, pitch_number,
            release_speed, release_spin_rate,
            pfx_x, pfx_z,
            release_extension,
            balls, strikes
        )

    write_csv(df, cache_file)
    df
}

gerrit_cole <- fetch_statcast("Gerrit Cole", pitcher_ids[["Gerrit Cole"]], "data/gerrit_cole.csv")
```

## Clean Data

`get_player_data()` filters a raw Statcast dataframe to a single pitcher, optionally subsets by pitch type and season, and restructures the result into the format expected by the HMM. Within each season, games are numbered sequentially by date. Within each game, pitches are sorted by `inning → at_bat_number → pitch_number` (the within-at-bat index) to recover true chronological order, then renumbered 1, 2, 3, ... as a single game-level sequence. The ordering columns are dropped; only the derived `pitch_number` is retained alongside the selected metrics.

```{r}
metrics_all <- c("release_speed", "release_spin_rate", "pfx_x", "pfx_z", "release_extension")

get_player_data <- function(data, name, pitch_type = NULL, season_filter = NULL, metrics = NULL) {
    if (is.null(metrics)) metrics <- metrics_all

    data %>%
        filter(player_name == name) %>%
        {
            if (!is.null(pitch_type)) filter(., pitch_type %in% !!pitch_type) else .
        } %>%
        {
            if (!is.null(season_filter)) filter(., season %in% season_filter) else .
        } %>%
        group_by(season, game_date) %>%
        arrange(inning, at_bat_number, pitch_number, .by_group = TRUE) %>%
        mutate(pitch_number = row_number()) %>%
        ungroup() %>%
        group_by(season) %>%
        mutate(game_number = dense_rank(game_date)) %>%
        ungroup() %>%
        select(season, player = player_name, game_number, pitch_number, all_of(metrics))
}

filtered_gerrit_cole <- get_player_data(gerrit_cole, "Cole, Gerrit", pitch_type = "FF", season_filter = max(gerrit_cole$season), metrics = "release_speed")
head(filtered_gerrit_cole)
```

## Fit Model

```{r}
# Use pre-2024 seasons to get empirical prior means for each state
historic_cole <- get_player_data(gerrit_cole, "Cole, Gerrit",
    pitch_type = "FF",
    season_filter = setdiff(unique(gerrit_cole$season), max(gerrit_cole$season)),
    metrics = "release_speed"
)

# Split at the 75th percentile: baseline (lower 75%) vs. hot (upper 25%)
q75 <- quantile(historic_cole$release_speed, 0.75)
baseline_pitches <- historic_cole$release_speed[historic_cole$release_speed <= q75]
hot_pitches <- historic_cole$release_speed[historic_cole$release_speed > q75]

prior_mu_mean <- c(mean(baseline_pitches), mean(hot_pitches))
cat(
    "Historic prior means — Baseline:", round(prior_mu_mean[1], 2),
    "  Hot:", round(prior_mu_mean[2], 2), "\n"
)
cat(
    "Group sizes — Baseline:", length(baseline_pitches),
    "  Hot:", length(hot_pitches), "\n"
)
```

```{r}
# Reshape cleaned data into list-of-games format expected by the HMM
Y <- split(filtered_gerrit_cole$release_speed, filtered_gerrit_cole$game_number)
Y <- lapply(Y, unname)
sigma <- sd(unlist(Y))

run_gibbs_chain <- function(Y, sigma, prior_nu, prior_pi, prior_mu_mean, prior_mu_var, init_nu, init_pi, init_mu, numsamp) {
    K <- length(init_mu)
    G <- length(Y)
    Tg <- sapply(Y, length)
    total_pitches <- sum(Tg)

    # Storage matrices
    mu.samp <- matrix(NA, nrow = numsamp, ncol = K)
    nu.samp <- matrix(NA, nrow = numsamp, ncol = K * K)
    pi.samp <- matrix(NA, nrow = numsamp, ncol = K)
    X.samp <- matrix(NA, nrow = numsamp, ncol = total_pitches)

    # Current parameter values
    curnu <- init_nu
    curpi <- init_pi
    curmu <- init_mu
    X <- lapply(Tg, function(t) rep(NA, t))

    for (iter in 1:numsamp) {
        #### Block 1: FFBS — sample hidden states per game ####
        for (g in 1:G) {
            alpha <- matrix(NA, nrow = Tg[g], ncol = K)

            # Forward pass
            for (i in 1:K) {
                alpha[1, i] <- curpi[i] * dnorm(Y[[g]][1], curmu[i], sigma)
            }
            if (Tg[g] > 1) {
                for (t in 2:Tg[g]) {
                    for (i in 1:K) {
                        alpha[t, i] <- 0
                        for (j in 1:K) {
                            alpha[t, i] <- alpha[t, i] +
                                alpha[t - 1, j] * curnu[j, i] * dnorm(Y[[g]][t], curmu[i], sigma)
                        }
                    }
                }
            }

            # Backward sampling
            prob_vec <- alpha[Tg[g], ] / sum(alpha[Tg[g], ])
            X[[g]][Tg[g]] <- sample(1:K, size = 1, prob = prob_vec)
            if (Tg[g] > 1) {
                for (t in (Tg[g] - 1):1) {
                    for (i in 1:K) {
                        prob_vec[i] <- alpha[t, i] * curnu[i, X[[g]][t + 1]]
                    }
                    prob_vec <- prob_vec / sum(prob_vec)
                    X[[g]][t] <- sample(1:K, size = 1, prob = prob_vec)
                }
            }
        }

        #### Block 2: Sample transition matrix ####
        Nmat <- matrix(0, nrow = K, ncol = K)
        for (g in 1:G) {
            for (t in 2:Tg[g]) {
                j <- X[[g]][t - 1]
                k <- X[[g]][t]
                Nmat[j, k] <- Nmat[j, k] + 1
            }
        }
        for (j in 1:K) {
            for (k in 1:K) {
                curnu[j, k] <- rgamma(1, shape = Nmat[j, k] + prior_nu[j, k], rate = 1)
            }
            curnu[j, ] <- curnu[j, ] / sum(curnu[j, ])
        }

        #### Block 3: Sample initial distribution ####
        init_count <- rep(0, K)
        for (g in 1:G) {
            init_count[X[[g]][1]] <- init_count[X[[g]][1]] + 1
        }
        for (j in 1:K) {
            curpi[j] <- rgamma(1, shape = init_count[j] + prior_pi[j], rate = 1)
        }
        curpi <- curpi / sum(curpi)

        #### Block 4: Sample emission means ####
        for (j in 1:K) {
            obs_j <- unlist(lapply(1:G, function(g) Y[[g]][X[[g]] == j]))
            n_j <- length(obs_j)
            if (n_j > 0) {
                Y_bar_j <- mean(obs_j)
                post_prec <- 1 / prior_mu_var + n_j / sigma^2
                post_mean <- (prior_mu_mean[j] / prior_mu_var + n_j * Y_bar_j / sigma^2) / post_prec
                curmu[j] <- rnorm(1, mean = post_mean, sd = sqrt(1 / post_prec))
            } else {
                curmu[j] <- rnorm(1, mean = prior_mu_mean[j], sd = sqrt(prior_mu_var))
            }
        }

        #### Store samples ####
        mu.samp[iter, ] <- curmu
        nu.samp[iter, ] <- as.vector(curnu)
        pi.samp[iter, ] <- curpi
        X.samp[iter, ] <- unlist(X)
    }

    list(mu = mu.samp, nu = nu.samp, pi = pi.samp, X = X.samp, game_lengths = Tg)
}

fix_label_switching <- function(mu_mat, nu_mat, pi_mat, X_mat) {
    n_iter <- nrow(mu_mat)
    for (i in 1:n_iter) {
        if (mu_mat[i, 1] > mu_mat[i, 2]) {
            # Swap mu
            temp_mu <- mu_mat[i, 1]
            mu_mat[i, 1] <- mu_mat[i, 2]
            mu_mat[i, 2] <- temp_mu

            # Swap pi
            temp_pi <- pi_mat[i, 1]
            pi_mat[i, 1] <- pi_mat[i, 2]
            pi_mat[i, 2] <- temp_pi

            # Swap nu (column-major vectorization: nu11, nu21, nu12, nu22)
            temp_nu11 <- nu_mat[i, 1]
            temp_nu21 <- nu_mat[i, 2]
            temp_nu12 <- nu_mat[i, 3]
            temp_nu22 <- nu_mat[i, 4]
            nu_mat[i, 1] <- temp_nu22
            nu_mat[i, 4] <- temp_nu11
            nu_mat[i, 2] <- temp_nu12
            nu_mat[i, 3] <- temp_nu21

            # Relabel X
            current_X <- X_mat[i, ]
            X_mat[i, current_X == 1] <- 2
            X_mat[i, current_X == 2] <- 1
        }
    }
    list(mu = mu_mat, nu = nu_mat, pi = pi_mat, X = X_mat)
}
```

```{r}
# Priors
K <- 2
prior_nu <- rbind(c(8, 1), c(1, 8))
prior_pi <- c(6, 2)
prior_mu_var <- 4 # SD = 2 mph
numsamp <- 10000
y_mean <- mean(unlist(Y))

cache_file <- "data/gerrit_cole_chains.rds"

if (file.exists(cache_file)) {
    message("Loading cached chains from ", cache_file)
    chains <- readRDS(cache_file)
    chain1 <- chains$chain1
    chain2 <- chains$chain2
} else {
    # Chain 1: starting values
    chain1 <- run_gibbs_chain(
        Y, sigma, prior_nu, prior_pi, prior_mu_mean, prior_mu_var,
        init_nu = rbind(c(0.75, 0.25), c(0.25, 0.75)),
        init_pi = c(0.5, 0.5),
        init_mu = c(y_mean - 1, y_mean + 1),
        numsamp = numsamp
    )

    # Chain 2: different starting values
    chain2 <- run_gibbs_chain(
        Y, sigma, prior_nu, prior_pi, prior_mu_mean, prior_mu_var,
        init_nu = rbind(c(0.5, 0.5), c(0.5, 0.5)),
        init_pi = c(0.75, 0.25),
        init_mu = c(y_mean - 2, y_mean + 2),
        numsamp = numsamp
    )

    saveRDS(list(chain1 = chain1, chain2 = chain2), cache_file)
    message("Chains saved to ", cache_file)
}

# Fix label switching in each chain
fixed1 <- fix_label_switching(chain1$mu, chain1$nu, chain1$pi, chain1$X)
fixed2 <- fix_label_switching(chain2$mu, chain2$nu, chain2$pi, chain2$X)
```

## MCMC Diagnostics

```{r, fig.height=9}
par(mfrow = c(4, 1))

# mu1 (Baseline state mean)
ymin <- min(fixed1$mu[, 1], fixed2$mu[, 1])
ymax <- max(fixed1$mu[, 1], fixed2$mu[, 1])
plot(1:numsamp, fixed1$mu[, 1],
    type = "l", col = 2, ylim = c(ymin, ymax),
    main = "Trace Plot: mu1 (Baseline)", ylab = "mu1", xlab = "Iteration"
)
lines(1:numsamp, fixed2$mu[, 1], col = 3)

# mu2 (Hot state mean)
ymin <- min(fixed1$mu[, 2], fixed2$mu[, 2])
ymax <- max(fixed1$mu[, 2], fixed2$mu[, 2])
plot(1:numsamp, fixed1$mu[, 2],
    type = "l", col = 2, ylim = c(ymin, ymax),
    main = "Trace Plot: mu2 (Hot)", ylab = "mu2", xlab = "Iteration"
)
lines(1:numsamp, fixed2$mu[, 2], col = 3)

# nu12 (Baseline -> Hot transition probability); column 3 of column-major as.vector(nu)
ymin <- min(fixed1$nu[, 3], fixed2$nu[, 3])
ymax <- max(fixed1$nu[, 3], fixed2$nu[, 3])
plot(1:numsamp, fixed1$nu[, 3],
    type = "l", col = 2, ylim = c(ymin, ymax),
    main = "Trace Plot: nu12 (Baseline -> Hot)", ylab = "nu12", xlab = "Iteration"
)
lines(1:numsamp, fixed2$nu[, 3], col = 3)

# nu21 (Hot -> Baseline transition probability); column 2 of column-major as.vector(nu)
ymin <- min(fixed1$nu[, 2], fixed2$nu[, 2])
ymax <- max(fixed1$nu[, 2], fixed2$nu[, 2])
plot(1:numsamp, fixed1$nu[, 2],
    type = "l", col = 2, ylim = c(ymin, ymax),
    main = "Trace Plot: nu21 (Hot -> Baseline)", ylab = "nu21", xlab = "Iteration"
)
lines(1:numsamp, fixed2$nu[, 2], col = 3)
```

## Posterior Summaries

## Posterior Predictive Checks

## Visualizations

---

# 4. Interpretation

## What Did We Learn?

## Limitations

## When Should We Use This Model?

## Sensitivity to Modeling Assumptions
